---
title: "Hypergate and C5.0"
author: "S. Granjeaud - CRCM"
date: "21 August 2018"
output:
  github_document:
    fig_width: 8
    fig_height: 6
    dev: jpeg
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Releases

| Version    | Changes
|------------|:---------------------------------------------------------------|
| 18/08/21   | hypergate and C5.0 1st release


| TODO       | Comment
|------------|:---------------------------------------------------------------|
|            | apply hypergate to mulitiple populations and compare to C5.0

## Introduction

Hypergate is a package that aims to characterize and find a signature of a selected set of events. The result could be interpreted as signature or a gating strategy. Hypergate is available at .

C5.0 is a decision tree algorithm that aims to find splits on original axes that allow to separate multiple groups of data points.

Let's compare their results.

Before add some helper functions.
```{r}

#' @title hgate_info
#' @description Extract information about a hypergate return: the channels of the phenotype, the sign of the channels, the sign of the comparison, the thresholds.
#' @param gate A hypergate object (produced by hypergate())
#' @return A data.frame with channel, sign, comp and threshold columns
#' @seealso \code{\link{hg_pheno}}, \code{\link{hg_rule}}
#' @examples
#' data(Samusik_01_subset)
#' xp=Samusik_01_subset$xp_src[,Samusik_01_subset$regular_channels]
#' gate_vector=Samusik_01_subset$labels
#' hg=hypergate(xp=xp,gate_vector=gate_vector,level=23,delta_add=0.01)
#' hgate_info(gate=hg)
#' hgate_pheno(gate=hg)
#' hgate_rule(gate=hg)
#' @export

hgate_info <- function(gate) {
  # retrieve threshold
  pars = gate$pars.history
  active_pars = gate$active_channels
  pars = pars[, active_pars, drop = FALSE]
  pars_order = apply(pars, 2, function(x) min(which(x != x[1])))
  pars = pars[, order(pars_order, decreasing = FALSE), drop = FALSE]
  pars = setNames(pars[nrow(pars), , drop = TRUE], colnames(pars))
  # get channel names
  channels = sub("_max", "", names(pars))
  channels = sub("_min", "", channels)
  # phenotype sign
  dir.sign = rep('+', length(pars))
  dir.sign[grep("_max", names(pars))] = '-'
  # comparison sign
  dir.comp = rep(' > ', length(pars))
  dir.comp[grep("_max", names(pars))] = ' <= '
# all together
  data.frame(
    channels, sign = dir.sign, comp = dir.comp, threshold = pars
  )
}

#' @title hgate_pheno
#' @description Build a human readable phenotype, i.e. a combination of channels and sign (+ or -) from a hypergate return.
#' @param gate A hypergate object (produced by hypergate())
#' @return A string representing the phenotype.
#' @seealso \code{\link{hg_rule}}, \code{\link{hg_info}}
#' @examples
#' ## See hgate_info
#' @export

hgate_pheno <- function(gate, collapse = ", ") {
  with(hgate_info(gate), paste0(channels, sign, collapse = collapse))
}

#' @title hgate_info
#' @description Extract information about a hypergate return: the channels of the phenotype, the sign of the channels, the sign of the comparison, the thresholds.
#' @param gate A hypergate object (produced by hypergate())
#' @return A data.frame with channel, sign, comp and threshold columns
#' @seealso \code{\link{hg_pheno}}, \code{\link{hg_rule}}
#' @examples
#' ## See hgate_info
#' @export

hgate_rule <- function(gate, collapse = ", ", digits = 2) {
  with(hgate_info(gate), paste0(channels, comp, round(threshold, digits), collapse = collapse))
}
```

## Hypergate example

### Rerun hypergate

```{r}
library(hypergate)

# Example data overview
data(Samusik_01_subset,package="hypergate")
head(Samusik_01_subset$labels, 10)
head(Samusik_01_subset$regular_channels, 10)
head(Samusik_01_subset$xp_src[,"label"], 10)

# Select a group of events
x = c(12.54, 8.08, 7.12, 12.12, 17.32, 20.62, 21.04, 20.83, 18.07, 15.2)
y = c(-10.61, -14.76, -18.55, -20.33, -21.16, -19.74, -14.4, -11.08, -10.02, -9.42)
pol = list(x = x, y = y)
gate_vector = sp::point.in.polygon(Samusik_01_subset$tsne[, 1], 
                                   Samusik_01_subset$tsne[, 2], pol$x, pol$y)
plot(Samusik_01_subset$tsne, pch = 16, cex = 0.5, 
     col = ifelse(gate_vector == 1, "firebrick3", "lightsteelblue"))
polygon(pol, lty = 2)
table(gate_vector)

# Define cluster colors (here there are 30 colors) from Nowicka et al F1000
color_clusters <- c("#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72",
  "#B17BA6", "#FF7F00", "#FDB462", "#E7298A", "#E78AC3",
  "#33A02C", "#B2DF8A", "#55A1B1", "#8DD3C7", "#A6761D",
  "#E6AB02", "#7570B3", "#BEAED4", "#666666", "#999999",
  "#aa8282", "#d4b7b7", "#8600bf", "#ba5ce3", "#808000",
  "#aeae5c", "#1e90ff", "#00bfff", "#56ff0d", "#ffff00")

plot(Samusik_01_subset$tsne, pch = 16, cex = 0.5, 
     col = color_clusters[as.factor(Samusik_01_subset$labels)])

# Compute hypergate
hg_output = hypergate(xp = Samusik_01_subset$xp_src[, Samusik_01_subset$regular_channels], gate_vector = gate_vector, level = 1, verbose = FALSE)
# hg_output

plot_gating_strategy(gate = hg_output, xp = Samusik_01_subset$xp_src[, 
    Samusik_01_subset$regular_channels], gate_vector = gate_vector, 
    level = 1, highlight = "firebrick3")
hgate_info(hg_output)

# Characterize the markers that target the group
gating_predicted = subset_matrix_hg(hg_output,
  Samusik_01_subset$xp_src[, Samusik_01_subset$regular_channels])
table(ifelse(gating_predicted, "Gated-in", "Gated-out"),
      ifelse(gate_vector == 1, "Events of interest", "Others"))
#
bm = boolmat(gate = hg_output,
  xp = Samusik_01_subset$xp_src[, Samusik_01_subset$regular_channels])
head(bm)
```

### C5.0 alternative

The typical model is below.
```{r}
library(C50)
model <- C5.0(Samusik_01_subset$xp_src[,Samusik_01_subset$regular_channels], as.factor(gate_vector))
summary(model)
```
C5.0 reports the same markers and order similarly. The importance of each marker is not the same as hypergate, but the message is the same. Out of 126 events selected on the tsne plot, 116 are correctly identified by the last rule, but 10 are missed.

We can also focuss on rules that gives an easier interpretation.
```{r}
model <- C5.0(Samusik_01_subset$xp_src[,Samusik_01_subset$regular_channels], as.factor(gate_vector), rules = TRUE)
summary(model)
```
Same summary, but result consists in rules.

Let's see the last one.
```{r}
# Rule 4: (116, lift 15.7)
# 	Ly6C <= 2.884321
# 	SiglecF > 2.152557
# 	cKit <= 1.794831
# 	->  class 1  [0.992]
in.rule = rep(TRUE, length(gate_vector))
in.rule = in.rule & Samusik_01_subset$xp_src[,"Ly6C"] <= 2.884321
in.rule = in.rule & Samusik_01_subset$xp_src[,"SiglecF"] > 2.152557
in.rule = in.rule & Samusik_01_subset$xp_src[,"cKit"]  <= 1.794831
table(in.rule, gate_vector)
# Columns are the truth, the groups we defined 
# Rows are the prediction the groups the algorithm defined
# The last rule retains 116 events, 116 belong to the group, 0 are out of the
# group, but 10 events of the group are missing
```


## Reducing markers

### Reoptimize strategy, without Ly6C

The proposed optimization without Ly6C in the signature.
```{r}
hg_output_polished = reoptimize_strategy(gate = hg_output, channels_subset = c("SiglecF_min", "cKit_max"), xp = Samusik_01_subset$xp_src[, Samusik_01_subset$regular_channels], gate_vector = gate_vector, level = 1)

gating_predicted_polished = subset_matrix_hg(hg_output_polished,
  Samusik_01_subset$xp_src[, Samusik_01_subset$regular_channels])
table(ifelse(gating_predicted_polished, "Gated-in", "Gated-out"),
      ifelse(gate_vector == 1, "Events of interest", "Others"))

plot_gating_strategy(gate = hg_output_polished, xp = Samusik_01_subset$xp_src[, 
    Samusik_01_subset$regular_channels], gate_vector = gate_vector, 
    level = 1, highlight = "firebrick3")
hgate_info(hg_output_polished)
```

Let's see if we remove the Ly6C marker from the signature.
```{r}
hg_output1 = hypergate(xp = Samusik_01_subset$xp_src[,setdiff(Samusik_01_subset$regular_channels, c("Ly6C"))], gate_vector = gate_vector, level = 1, verbose = FALSE)
gating_predicted1 = subset_matrix_hg(hg_output1,
  Samusik_01_subset$xp_src[, setdiff(Samusik_01_subset$regular_channels, c("Ly6C"))])
table(ifelse(gating_predicted, "Gated-in", "Gated-out"),
      ifelse(gate_vector == 1, "Events of interest", "Others"))

plot_gating_strategy(gate = hg_output1, xp = Samusik_01_subset$xp_src[, 
    Samusik_01_subset$regular_channels], gate_vector = gate_vector, 
    level = 1, highlight = "firebrick3")
hgate_info(hg_output1)
```
The result is nearly the same.

Let's see if we remove the SiglecF marker from the signature.
```{r}
hg_output_polished2 = reoptimize_strategy(gate = hg_output, channels_subset = c("SiglecF_min"), xp = Samusik_01_subset$xp_src[, Samusik_01_subset$regular_channels], 
    gate_vector = gate_vector, level = 1)

plot_gating_strategy(gate = hg_output_polished2, xp = Samusik_01_subset$xp_src[, 
    Samusik_01_subset$regular_channels], gate_vector = gate_vector, 
    level = 1, highlight = "firebrick3")

gating_predicted_polished2 = subset_matrix_hg(hg_output_polished2,
  Samusik_01_subset$xp_src[, Samusik_01_subset$regular_channels])
table(ifelse(gating_predicted_polished2, "Gated-in", "Gated-out"),
      ifelse(gate_vector == 1, "Events of interest", "Others"))
hgate_info(hg_output_polished2)
```
It still perform well using other markers.


### C5.0 alternative

Maybe we can prune the tree at a higher level, but let's remove a marker and redo computation.

Let's remove Ly6C from markers.
```{r}
# without Ly6C
model_reoptimzed <- C5.0(Samusik_01_subset$xp_src[,setdiff(Samusik_01_subset$regular_channels, c("Ly6C"))], as.factor(gate_vector), rules = TRUE)
summary(model_reoptimzed)

# Rule 4: (114/1, lift 15.6)
# F480 > 1.084815
# SiglecF > 2.152557
# cKit <= 1.794831
# ->  class 1  [0.983]
# While 13 events are missing, nearly all events are correctly identified
in.rule = rep(TRUE, length(gate_vector))
in.rule = in.rule & Samusik_01_subset$xp_src[,"F480"] > 1.084815
in.rule = in.rule & Samusik_01_subset$xp_src[,"SiglecF"] > 2.152557
in.rule = in.rule & Samusik_01_subset$xp_src[,"cKit"] <= 1.794831
table(in.rule, gate_vector)

# Rule 5: (114/1, lift 15.6)
# SiglecF > 2.152557
# cKit <= 1.794831
# CD43 <= 2.243915
# ->  class 1  [0.983]
# same result with this slightly different rule
in.rule = rep(TRUE, length(gate_vector))
in.rule = in.rule & Samusik_01_subset$xp_src[,"SiglecF"] > 2.152557
in.rule = in.rule & Samusik_01_subset$xp_src[,"cKit"] <= 1.794831
in.rule = in.rule & Samusik_01_subset$xp_src[,"CD43"] <= 2.243915
table(in.rule, gate_vector)
```

Let's remove SiglecF from markers.
```{r}
# without SiglecF
model_reoptimzed2 <- C5.0(Samusik_01_subset$xp_src[,setdiff(Samusik_01_subset$regular_channels, c("SiglecF"))], as.factor(gate_vector), rules = TRUE)
summary(model_reoptimzed2)
# no simple rule
# Rule 10: (79, lift 15.7)
# 	F480 > 1.157256
# 	Ly6C <= 1.869782
# 	Foxp3 > 0.07933632
# 	cKit <= 1.819595
# 	CD43 <= 2.478766
# 	CD44 > 3.698416
# 	->  class 1  [0.988]
```


## Expand the number of populations

The graphical selection of hypergate corresponds mainly to pop 8, ie Eosinophils.
```{r}
# Identify the targetted cluster among annotated events
table(gate_vector, Samusik_01_subset$labels)

pops = c("B-cell Frac A-C (pro-B cells)", "Basophils", "CD4 T cells", "CD8 T cells", "Classical Monocytes", "CLP", "CMP", "Eosinophils", "gd T cells", "GMP", "HSC", "IgD- IgMpos B cells", "IgDpos IgMpos B cells", "IgM- IgD- B-cells", "Intermediate Monocytes", "Macrophages", "mDCs", "MEP", "MPP", "NK cells", "NKT cells", "Non-Classical Monocytes", "pDCs", "Plasma Cells")

# Check the amount of populations in the sample
table(Samusik_01_subset$labels)

# Let select some of the clusters
gate_vector_multi = Samusik_01_subset$labels
gate_vector_multi[!Samusik_01_subset$labels %in% c(5, 8, 12, 13, 15, 23)] = 0
table(gate_vector_multi)
```


### C5.0

Let's apply C5.0 to all those groups at once.
```{r, results="asis"}
model <- C5.0(Samusik_01_subset$xp_src[,Samusik_01_subset$regular_channels], as.factor(gate_vector_multi), rules = TRUE, control = C5.0Control(winnow = TRUE, noGlobalPruning = TRUE))
C50_head_tail <- function(model, head = 20, tail = 30) {
  model_str <- strsplit(model$output, "\\n")[[1]]
  if (length(model_str) < head + tail) return(model_str)
  c(model_str[1:head], "", "...truncated result...", "", model_str[length(model_str)-(tail:1)])
}
C50_head_tail(model)
```
C5.0 is performing very well.

Let's find best rules using F scores.
```{r}
# get rules of the model
c50.rules = strsplit(model$rules, "\\n")[[1]]
# parse rules attributes
rules = c("conds,cover,TP,lift,class")
for (i in grep("conds", c50.rules)) {
  rules = c(rules, sub("conds=\"(\\d+)\" cover=\"(\\d+)\" ok=\"(\\d+)\" lift=\"([0-9.]+)\" class=\"(\\d+)\"", "\\1,\\2,\\3,\\4,\\5", c50.rules[i]))
}
rules = read.table(text=paste(rules, collapse = "\n"), sep = ",", header = TRUE)
targets = as.data.frame(table(gate_vector_multi))
colnames(targets) = c("class", "Count")
rules = merge(rules, targets)
rules = within(rules, {
  FP = cover - TP
  FN = Count - TP
  recall = TP/(TP + FN)
  purity = TP/(TP + FP)
  beta = 1
  Fscore = ifelse(is.nan(purity) | is.nan(recall), 0, 
                  (1 + beta^2) * (recall * purity)/(recall + beta^2 * purity))
  rm(beta)
})
head(rules)

# find best rule for each class
rules.best.idx = sapply(unique(rules$class), function(cl) { c50.rules = rules[rules$class == cl, , drop = FALSE]; rownames(c50.rules[which.max(c50.rules$Fscore), , drop = FALSE])})
rules.best = rules[rules.best.idx,]
# Compare True Positive (TP) and Count (from manual annotation)
library(knitr)
kable(rules.best)
```

Let's extract rule defintions.
```{r}
# parse rules
rule.defs = c("rule,type,att,cut,result")
rule.id = 0
for (i in seq(c50.rules)) {
  if (substr(c50.rules[i], 1, 6) == "conds=") {
    rule.id = rule.id + 1
  }
  if (substr(c50.rules[i], 1, 5) == "type=") {
    rule.defs = c(rule.defs, paste0(c(rule.id, sub("type=\"(\\d+)\" att=\"(.+)\" cut=\"([0-9.-]+)\" result=\"([<>])\"", "\\1,\\2,\\3,\\4", c50.rules[i])), collapse = ","))
  }
}
rule.defs = read.table(text=paste(rule.defs, collapse = "\n"), sep = ",", header = TRUE)

# Helpers for C5.0
C50_info <- function(gate) {
  with(gate, data.frame(
    channels = att, sign = ifelse(result == "<", "-", "+"), comp = result, threshold = cut
  ))
}
C50_pheno <- function(gate, collapse = ", ") {
  with(C50_info(gate), paste0(channels, sign, collapse = collapse))
}

C50_rule <- function(gate, collapse = ", ", digits = 2) {
  with(C50_info(gate), paste0(channels, comp, round(threshold, digits), collapse = collapse))
}
```

Extract simplified population definitions for biologist.
```{r}
# Extract simplified population definitions for biologist
kable(sapply(rownames(rules.best), function(i) C50_pheno(subset(rule.defs, rule == i))))
kable(sapply(rownames(rules.best), function(i) C50_rule(subset(rule.defs, rule == i))))

# cbind
kable(rules.best)
kable(cbind(pheno = sapply(rownames(rules.best), function(i) C50_pheno(subset(rule.defs, rule == i))), pop = c("Others", pops[rules.best$class])))
```
